---
title: "Example Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.fullwidth = TRUE,
  fig.width = 8,
  fig.height = 5
)
```

```{r setup}
library(nlme)
library(ggplot2)
```

In this vignette, we demonstrate how to conduct a linear mixed model using the `nlme` package.

Here, we use data featured in TidyTuesday Podcast Episode 16: College Tuition](https://www.tidytuesday.com/16). This is a longitudinal data set of tuition fees from the National Center of Education Statistics (NCES). Instructions for downloading the data, as well as the data dictionary are available [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-03-10). We will use the `tuition_income.csv` file.

We will aim to answer the question: What is the mean change in on-campus tuition prices from 2014 to 2018 across U.S. universities? We will also answer two sub-questions: (1) What is the mean on-campus tuition price every year from 2014 to 2018 per state, and (2) What is the mean five-year change in on-campus tuition price for every state?

We first read the data in:

```{r read_data}
# Store the name of the csv on TidyTuesday github
tuition_income_url <- "tuition_income.csv"

# Test if a directory named data exists locally (will create if it doesn't exist)
ifelse(!dir.exists("data"), dir.create("data"),
       "Directory is already created")

# Read the data only once
tuition_income_path <- "data/tuition_income.Rdata"

if (!file.exists(tuition_income_path)) {
  tuition_income <-
    paste0("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/", tuition_income_url) |>
    readr::read_csv() |>
    janitor::clean_names()
  save(tuition_income, file = tuition_income_path)
} else {
  load(tuition_income_path)
}
```

We first subset our dataset to retrieve only data for the years from 2014 to 2016 and on campus tuition prices. Since the data set is further stratified by the net price by income bracket, we retrieve only the first observation within each school.

```{r dataclean}
# Filter data from 2014 to 2018 and only on-campus tuition
tuition_income_1418_oc <- tuition_income |>
  dplyr::filter(year %in% c(2014:2018) & campus == "On Campus") |>
  dplyr::group_by(name, year) |>
  dplyr::filter(dplyr::row_number() == 1L) |>
  dplyr::ungroup() |>
  dplyr::select(name, state, total_price, year)

tuition_income_1418_oc
```

We answer the first question: What is the mean on-campus tuition price every year from 2014 to 2018 per state?

```{r mean_yearstate}
# Create a data set per year
tuition_income_1418_oc_byyear <- tuition_income_1418_oc |> 
  split(tuition_income_1418_oc$year)

# Create a function that calculates the mean tuition by state and stores in a data frame
meantuition_state <- function(x) {
  # x: index of list in tuition_income_1418_oc_byyear
  means <- tuition_income_1418_oc_byyear[[x]] |>
    split(tuition_income_1418_oc_byyear[[x]]$state) |>
    purrr::map_dfr(.f = ~mean(.x$total_price)) |>
    t()
  data.frame(state = rownames(means),
             meantuition = means[,1],
             year = names(tuition_income_1418_oc_byyear[x]))
}

# Apply the function to each year and store in one data frame
meantuition_yearstate <- purrr::map_dfr(1:5, .f = meantuition_state)
```

We then present the results in a choropleth map, by year:

```{r map1, fig.width = 8, fig.height = 4}
# Match the state abbreviation with the state name
meantuition_yearstate <- meantuition_yearstate |>
  dplyr::mutate(region = tolower(state.name[match(state, state.abb)]))

# Retrieve the states map data and merge with mean tuition data
states_map <- ggplot2::map_data("state")
meantuition_yearstate_map <- dplyr::left_join(states_map, meantuition_yearstate, by = "region") |>
  dplyr::filter(!is.na(year))

# Create the choropleth map
ggplot(meantuition_yearstate_map, aes(long, lat, group = group)) + 
  geom_polygon(aes(fill = meantuition), color = "white") +
  scale_fill_viridis_c() + 
  facet_wrap(~year, ncol = 3) +
  labs(fill = "Mean tuition",
       title = "Mean tuition cost was highest in the Northeast from 2014-2018",
       subtitle = "Mean tuition did not change much across states",
       caption = "Tuitiion cost data from National Center of Education Statistics (NCES)") +
  theme_void() +
  theme(legend.position = c(0.9,0.1), legend.justification = c(0.9,0.1))
```
We see that the highest tuition costs were in the Northeast: Washington DC, Rhode Island, and Massachusetts had average four-year college tuition costs at more than 50,000 USD a year. On the other hand, Wyoming, Mississippi, and North Dakota had the lowest tuition costs at less than 15,000 USD a year.

We then answer the second question: What is the mean five-year change in on-campus tuition price for every state?

First, we visualize the tuition cost every year for each school using a spaghetti plot, highlighting the highest three and lowest three states we mentioned earlier.

```{r spagplot}
tuition_income_1418_oc_hh <- tuition_income_1418_oc |>
  dplyr::mutate(highlight = ifelse(state %in% c("DC", "RI", "MA", "ND", "MS", "WY"), state, "")) |>
  dplyr::arrange(highlight)


ggplot(tuition_income_1418_oc_hh, 
       aes(group = name, x = year, y = total_price, 
           color = as.factor(highlight), alpha = as.factor(highlight))) +
  geom_line() +
  scale_color_manual(breaks = c("", "DC", "RI", "MA", "ND", "MS", "WY"),
                     values = c("gray", "#4040DB", "#7D52F2", "#9340DB",
                                "#DE5445", "#F57C59", "#DE7845")) +
  scale_alpha_manual(breaks = c("", "DC", "RI", "MA", "ND", "MS", "WY"),
                     values = c(0.2, 1, 1, 1, 1, 1, 1)) +
  labs(color = "", alpha = "", x = "Year", y = "On-campus tuition cost",
       title = "Tuition costs were increasing across all schools in the United States",
       subtitle = "Top 3 highest and lowest cost states remained in their ranks across time",
       caption = "Tuition cost data from National Center of Education Statistics (NCES)") +
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12, face = "bold"),
        axis.line = element_line(linewidth = 0.5),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.title = element_blank()
)
```
We then run two candidate models: a random intercept model with year and state as covariates, and another adding year*state interaction terms. We use the `lme` function to run the models, and the `anova.lme` function to compare the models using a likelihood ratio test from the `nlme` package. 

```{r rintmod}
# Random intercept model without interaction term
model_noint <- lme(fixed = total_price ~ year + state,
                 random = ~ 1 | name,
                 data = tuition_income_1418_oc,
                 method = "ML") # ML to allow likelihood ratio test instead of default REML

# Random intercept model with interaction term
model_int <- lme(fixed = total_price ~ year*state,
                 random = ~ 1 | name,
                 data = tuition_income_1418_oc,
                 method = "ML")

# Likelihood ratio test
anova.lme(model_int, model_noint)
```

We see that the model with year*state interaction terms significantly improve model fit, so we use this model to estimate the mean annual change of tuition cost for each state.

```{r meanchangestate}
# Retrieve the coefficients
year_coef <- model_int$coefficients$fixed[2]

# Retrieve the state coefficients
# Note that AK is the "comparison" state, so their coefficient is just 0
states_coef <- c(0, model_int$coefficients$fixed[53:102])

# Retrieve the state coefficient names
states_names <- levels(factor(tuition_income_1418_oc$state))

# Calculate the annual change in tuition cost per year per state
states_change <- states_coef |> purrr::map_dbl(function(x) x + year_coef)

# Assemble in a data frame
states_change_df <- data.frame(state = states_names,
                               meanchange = states_change)
```

We then present the results using a bar graph, ranked by mean annual change:

```{r plotmeanchange, fig.height = 10}
ggplot(states_change_df, aes(x = meanchange, y = reorder(state, meanchange))) +
  geom_bar(stat = "identity", color = "gray", alpha = 0.2) +
  labs(color = "", alpha = "", x = "Mean annual increase in tuition cost (USD)", y = "",
       title = "Mean annual increase in tuition cost ranged from 300 to 1,700 USD across states",
       subtitle = "Top 3 highest states were Rhode Island, Massachusetts, and Washington DC",
       caption = "Tuition cost data from National Center of Education Statistics (NCES)") +
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12, face = "bold"),
        axis.line = element_line(linewidth = 0.5),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.title = element_blank()
)
```

# Summary

Based from data on the National Center of Education Statistics (NCES) from 2014 to 2018, all schools increased their on-campus tuition cost for a four-year undergraduate degree. The highest costs, as well as the highest mean annual change in cost, were in Rhode Island, Massachusetts, and Washington DC.

# Functions used

* `nlme`: `lme`, `anova.lme`
* `dplyr`: `arrange`, `filter`, `group_by`, `left_join`, `mutate`, `select`, `ungroup`
* `purrr`: `map_dfr`, `map_dbl`
* `ggplot2` geoms: `geom_polygon`, `geom_line`
* `readr`: `read_csv`
* `janitor`: `clean_names`